KiCS2: A New Compiler from Curry to Haskell

### 2011-04-20 papers/wflp2011/paper.tex
Paper refinements

\begin{abstract}
In this paper we present our first results of a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic result values of an expression in a single data structure.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

In this paper we describe an implementation that is based
on the latter principle. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs based on the ideas shown
in \cite{BrasselFischer08IFL}.
The difficulty of such an implementation is the fact
that non-deterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
but the outcome of any operation could be potentially non-deterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
non-deterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of the Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation, called KiCS2, is not based on unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate non-deterministic computations
(which are not described in detail in this paper due to lack of space).

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{lstlisting}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{lstlisting}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics is the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\label{sec:no-logic-vars}
First of fall, we assume that our programs
\emph{do not contain logic variables}.
This assumption can be made since it has been shown
\cite{AntoyHanus06ICLP} that logic variables can be replaced
by non-deterministic ``generators,''
i.e., operations that evaluate to all possible values of the type of
the logic variable. For instance, a Boolean logic variable
can be replaced by the generator \code{aBool} defined above.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
%                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules where only one constructor occurs in the same argument
of all rules.\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

As mentioned above, our implementation is based on the explicit
representation of non-deterministic results in a data structure.
This could be easily achieved by adding a constructor
to each data type to represent a choice between two values.
For instance, one can redefine the data type for Boolean values
as follows:
\begin{haskell}
  data Bool = False | True | Choice Bool Bool
\end{haskell}
Thus, we can implement the non-deterministic operation \code{aBool}
defined in Section~\ref{ex:aBool} as follows:
\begin{haskell}
  aBool = Choice True False
\end{haskell}
If operations can deliver non-deterministic values,
we have to extend the rules for operations defined by
pattern matching so that they do not fail on non-deterministic argument
values. Instead, they move the non-deterministic choice one level above,
i.e., a choice in some argument leads to a choice in any result
of this operation (this is also called a ``pull-tab'' step in
\cite{AlqaddoumiAntoyFischerReck10}). For instance,
the rules of the uniform operation \code{xor} shown above
are extended as follows:
%
\begin{curry}
  xor True           x = xor' x
  xor False          x = x
  xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with different
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:\footnote{%
Consequently, integers cannot be used for \code{IDSupply}s. Nevertheless,
\code{IDSupply}s can be implemented as an infinite tree containing
all memory cells required in a computation that are created
on demand with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
\hline
System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\
PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\
MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\
MonC    &   48.15  &      916.61   &   n/a &    n/a \\
\hline
\end{tabular}
\caption{Benchmarks: non-deterministic functional logic programs}
 \label{fig:bench-nondet}
\end{figure}
%
To evaluate the efficiency of non-deterministic computations
(Fig.~\ref{fig:bench-nondet}),
we sort a list containing 15 elements by enumerating all permutations
and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
for Peano numbers),
compute the last element \code{x}
of a list \code{xs} containing 100,000 elements
by solving the equation \ccode{ys++[x] =:= xs}
(the implementation of unification and variable bindings
require some additional machinery that is sketched in
Section~\ref{sec:unification}),
and match a regular expression in a string of length 200,000
following the non-deterministic specification of \code{grep} shown
in \cite{AntoyHanus10CACM}.
The results show that our high-level implementation is not far
from the efficiency of MCC,
and it is superior to PAKCS which exploits Prolog features
like backtracking, logic variables and unification for these benchmarks.

Since our implementation represents non-deterministic values
as Haskell data structures, we get, in contrast to most
other implementations of Curry, one interesting improvement for free:
deterministic subcomputations are shared even if they occur
in different non-deterministic computations.
To show this effect of our implementation, consider the
non-deterministic sort operation \code{psort} (permutation sort)
and the infinite list of all prime numbers \code{primes},
as used in the previous benchmarks, and the following definitions:
\begin{curry}
 goal1 = [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
 goal2 = psort [7949,7937,7933,7927]
 goal3 = psort [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
\end{curry}
In principle, one would expect that the sum of the execution times
of \code{goal1} and \code{goal2} is equal to the time to execute
\code{goal3}. However, implementations based on backtracking
evaluate the primes occurring in \code{goal3} multiple times,
as can be seen by the run times for PAKCS and MCC shown
in Fig~\ref{fig:bench-sharing-over-nondet}.
%
\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
System  & \code{goal1} & \code{goal2} & \code{goal3} \\\hline
KiCS2   &        0.34  &        0.00  &        0.34  \\
PAKCS   &       14.90  &        0.02  &      153.65  \\
MCC     &        0.33  &        0.00  &        3.46  \\
\hline
\end{tabular}
\caption{Benchmarks: sharing over non-determinism}
 \label{fig:bench-sharing-over-nondet}
\end{figure}

>>>>

\begin{abstract}
In this paper we present our first results of a new system % first steps towards ?
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic values in the respective data structure.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

In this paper we describe an implementation that is based
on the latter principle. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs based on the ideas shown
in \cite{BrasselFischer08IFL}.
The difficulty of such an implementation is the fact
that non-deterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
as the outcome of any operation could be potentially non-deterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
non-deterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of the Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation, called KiCS2, avoids such unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate non-deterministic computations
(which are not described in detail in this paper due to lack of space).

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{lstlisting}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{lstlisting}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\label{sec:no-logic-vars}
First of all, we assume that our programs
\emph{do not contain logic variables}.
This assumption can be made since it has been shown
\cite{AntoyHanus06ICLP} that logic variables can be replaced
by non-deterministic ``generators,''
i.e., operations that evaluate to all possible values of the type of
the logic variable. For instance, a Boolean logic variable
can be replaced by the generator \code{aBool} defined above.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
%                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in each rule, and
in the same argument for all rules.
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

As mentioned above, our implementation is based on the explicit
representation of non-deterministic results in a data structure.
This can easily be achieved by adding a constructor
to each data type to represent a choice between two values.
For instance, one can redefine the data type for Boolean values
as follows:
\begin{haskell}
  data Bool = False | True | Choice Bool Bool
\end{haskell}
Thus, we can implement the non-deterministic operation \code{aBool}
defined in Section~\ref{ex:aBool} as:
\begin{haskell}
  aBool = Choice True False
\end{haskell}
If operations can deliver non-deterministic values,
we have to extend the rules for operations defined by
pattern matching so that they do not fail on non-deterministic argument
values. Instead, they move the non-deterministic choice one level above,
i.e., a choice in some argument leads to a choice in any result
of this operation (this is also called a ``pull-tab'' step in
\cite{AlqaddoumiAntoyFischerReck10}). For instance,
the rules of the uniform operation \code{xor} shown above
are extended as follows:
%
\begin{curry}
  xor True           x = xor' x
  xor False          x = x
  xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with different
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:\footnote{%
Consequently, integers cannot be used for \code{IDSupply}s. Instead,
\code{IDSupply}s can be implemented as an infinite tree containing
all memory cells required in a computation that are created
on demand with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
\hline
System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
KiCS2HO &    2.83  &        3.68   &  0.14 &   0.49 \\
PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\
MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\
MonC    &   48.15  &      916.61   &   n/a &    n/a \\
\hline
\end{tabular}
\caption{Benchmarks: non-deterministic functional logic programs}
 \label{fig:bench-nondet}
\end{figure}
%
To evaluate the efficiency of non-deterministic computations
(Fig.~\ref{fig:bench-nondet}),
we sort a list containing 15 elements by enumerating all permutations
and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
for Peano numbers),
compute the last element \code{x}
of a list \code{xs} containing 100,000 elements
by solving the equation \ccode{ys++[x] =:= xs}
(the implementation of unification and variable bindings
require some additional machinery that is sketched in
Section~\ref{sec:unification}),
and match a regular expression in a string of length 200,000
following the non-deterministic specification of \code{grep} shown
in \cite{AntoyHanus10CACM}.
The results show that our high-level implementation is not far
from the efficiency of MCC,
and it is superior to PAKCS which exploits Prolog features
like backtracking, logic variables and unification for these benchmarks.

Since our implementation represents non-deterministic values
as Haskell data structures, we get, in contrast to most
other implementations of Curry, one interesting improvement for free:
deterministic subcomputations are shared even if they occur
in different non-deterministic computations.
To show this effect of our implementation, consider the
non-deterministic sort operation \code{psort} (permutation sort)
and the infinite list of all prime numbers \code{primes},
as used in the previous benchmarks, and the following definitions:
\begin{curry}
 goal1 = [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
 goal2 = psort [7949,7937,7933,7927]
 goal3 = psort [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
\end{curry}
In principle, one would expect that the sum of the execution times
of \code{goal1} and \code{goal2} is equal to the time to execute
\code{goal3}. However, implementations based on backtracking
evaluate the primes occurring in \code{goal3} multiple times,
as can be seen by the run times for PAKCS and MCC shown
in Fig~\ref{fig:bench-sharing-over-nondet}.
%
\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
System  & \code{goal1} & \code{goal2} & \code{goal3} \\\hline
KiCS2HO &        0.34  &        0.00  &        0.34  \\
PAKCS   &       14.90  &        0.02  &      153.65  \\
MCC     &        0.33  &        0.00  &        3.46  \\
\hline
\end{tabular}
\caption{Benchmarks: sharing over non-determinism}
 \label{fig:bench-sharing-over-nondet}
\end{figure}
